---
title: "AgriCerrado Calibration Report"
author: "João Marcolin"
date: "08/05/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width='800px')

library(magrittr)
require(tidyverse)
library(knitr)
library(pander)
library(reshape2)
library(kableExtra)
library(viridis)
library(patchwork)
library(scales)
theme_set(theme_grey())

## define auxiliary function to save figures
# picks a folder to save the figures (if necessary, create the folder)
if (!dir.exists("figures")) dir.create("figures")
# auxiliary function for saving figures
save_fig <- function(plot, name, width = 8, height = 6, dpi = 500) {
  ggsave(paste0("figures/", name, ".png"), plot = plot, width = width, height = height, dpi = dpi)
}

```

```{r load, include=FALSE}
# load inputs
df_brazil_mun  <- readr::read_csv(
  "data_outputs/2_brazil_mun/df_brazil_mun.csv",
  col_types = cols(
    .default               = col_character(),
    legal_amazon           = col_logical(),
    mun_area_km2           = col_number(),
    mun_protected_area_km2 = col_number(),
    FAO_soy_yield          = col_number()
    )
  )

df_census_state  <- readr::read_csv(
  "data_outputs/3_census_data/df_census_state.csv",
  col_types = cols(
    .default      = col_character(),
    year          = col_number(),
    farm_area_ha  = col_number(),
    n_farms       = col_number(),
    activity_area = col_number(),
    workers       = col_number()
    )
  )

df_census_mun <- readr::read_csv(
  "data_outputs/3_census_data/df_census_mun.csv",
  col_types = cols(
    .default      = col_character(),
    legal_amazon  = col_logical(),
    year          = col_number(),
    farm_area_ha  = col_number(),
    n_farms       = col_number(),
    activity_area = col_number(),
    workers       = col_number()
    )
  )

df_yearly_mun    <- readr::read_csv(
  "data_outputs/4_yearly_data/df_yearly_mun.csv",
  col_types = cols(
    .default = col_character(),
    legal_amazon   = col_logical(),
    year           = col_number(),
    pam_area_ha    = col_number(),
    pam_output_ton = col_number(),
    ppm_herd       = col_number(),
    mb_natcover_ha = col_number(),
    mb_nonveg_ha   = col_number(),
    mb_notobs_ha   = col_number(),
    mb_pasture_ha  = col_number(),
    mb_soybean_ha  = col_number(),
    mb_tcrop_ha    = col_number(),
    mb_forestry_ha = col_number(),
    mb_pcrop_ha    = col_number()
    )
  )

df_yearly_prices <- readr::read_csv(
  "data_outputs/4_yearly_data/df_prices.csv",
  col_types = cols(
    .default = col_number()
    )
  )

# load municipalities shapefiles from 2024
cities_sf <- sf::st_read("data_outputs/1_cell_grid/1_reproject/cities_EPSG5880.shp", quiet = TRUE) %>%
  dplyr::filter(state %in% report_states)

```

# Output prices, minimum wage and interest rate

```{r yearly_prices, echo=FALSE}

# auxiliary function to plot yearly values (line plot)
plot_var <- function(df, var, var_name, title) {
  ggplot(df, aes(x=year, y=.data[[var]])) +
    geom_line() +
    labs(title=title, x="Year", y=var_name) +
    theme_minimal()
}

# fix NAs in price_C and price_S
df_yearly_prices$price_C[1:10] <- NA
df_yearly_prices$price_S[1:10] <- NA

# plot
plot_interest <- plot_var(df_yearly_prices, "interest", "interest", "Cumulative yearly interest rate") #+
                   #geom_smooth(method = "loess", se=FALSE) # add trend line
plot_wage     <- plot_var(df_yearly_prices, "wage",     "min. wage", "National minimum wage") #+
                   #geom_smooth(method = "loess", se=FALSE) # add trend line
plot_price_C  <- plot_var(df_yearly_prices, "price_C",  "BRL per ton", "Beef price (base=2024)") #+
                   #geom_smooth(method = "loess", se=FALSE) # add trend line
plot_price_S  <- plot_var(df_yearly_prices, "price_S",  "BRL per ton", "Soybean price (base=2024)") #+
                   #geom_smooth(method = "loess", se=FALSE) # add trend line

# display plots
(plot_interest | plot_wage) / (plot_price_C | plot_price_S)

# save plots
save_fig(plot_interest, "Brazil_interest")
save_fig(plot_wage,     "Brazil_wage")
save_fig(plot_price_C,  "Brazil_price_C")
save_fig(plot_price_S,  "Brazil_price_S")

# fill NAs back
df_yearly_prices$price_C[1:10] <- df_yearly_prices$price_C[11]
df_yearly_prices$price_S[1:10] <- df_yearly_prices$price_S[11]

rm(plot_interest, plot_wage, plot_price_C, plot_price_S, df_yearly_prices2)
```

# Beef and Soybean yield
## National and State yields over time

```{r yearly_agg_yield, echo=FALSE}
# summarise at the state level
yearly_state_data <- df_yearly_mun %>%
  dplyr::select(
    year, state,
    mb_pasture_ha, mb_soybean_ha,
    ppm_herd, pam_output_ton, pam_area_ha,
    ) %>%
  dplyr::group_by(
    state, year
    ) %>%
  dplyr::summarise(
    across(everything(), ~sum(.x,na.rm=TRUE))
  ) %>%
  dplyr::mutate(
    yield_C = ppm_herd/mb_pasture_ha,
    yield_S = pam_output_ton/pam_area_ha
  )

# summarise at the national level
yearly_brazil_data <- df_yearly_mun %>%
  dplyr::transmute(
    year, state = "Brazil",
    mb_pasture_ha, mb_soybean_ha,
    ppm_herd, pam_output_ton, pam_area_ha,
    ) %>%
  dplyr::group_by(
    state, year
    ) %>%
  dplyr::summarise(
    across(everything(), ~sum(.x,na.rm=TRUE))
  ) %>%
  dplyr::mutate(
    yield_C = ppm_herd/mb_pasture_ha,
    yield_S = pam_output_ton/pam_area_ha
  )
# fix yield_S_mb: it's way too high in 1985
yearly_brazil_data$yield_S_mb[1] <- yearly_brazil_data$yield_S_mb[2]

# plot country-level data
plot_yield_C <- plot_var(yearly_brazil_data, "yield_C", "tons per ha.", "Cattle yield (head/ha., Brazil)") +
  geom_smooth(method = "loess", se=FALSE) # add trend line
plot_yield_S <- plot_var(yearly_brazil_data, "yield_S", "tons per ha.", "Soybean yield (tons/ha., Brazil)") +
  geom_smooth(method = "loess", se=FALSE) # add trend line

# plot state-level data
# plot yield_C
plot_yield_C_state <- yearly_state_data %>%
  dplyr::filter(state %in% report_states) %>%
  ggplot(aes(x=year, y=yield_C, color=state)) +
  geom_line() +
  labs(title="Cattle yield (head/ha.)", x="Year", y="tons/ha", color="State") +
  theme_minimal()

# plot yield_S
plot_yield_S_state <- yearly_state_data %>%
  dplyr::filter(state %in% report_states) %>%
  ggplot(aes(x=year, y=yield_S, color=state)) +
  geom_line() +
  labs(title="Soybean yield (tons/ha.)", x="Year", y="tons/ha", color="State") +
  theme_minimal()

(plot_yield_C | plot_yield_S)
(plot_yield_C_state | plot_yield_S_state)

save_fig(plot_yield_C, "Brazil_yield_C")
save_fig(plot_yield_S, "Brazil_yield_S")

save_fig(plot_yield_C_state, "states_yield_C")
save_fig(plot_yield_S_state, "states_yield_S")

rm(plot_yield_C, plot_yield_S,
   plot_yield_C_state, plot_yield_S_state,
   yearly_brazil_data, yearly_state_data)


```

## Distribution of municipality-level soybean yield

```{r mun_yield_S_distrib, echo=FALSE}
# auxiliary function to summarise by municipality group
summarise_mun_yield_f <- function(yr) {
  # define grouping column
  group_col <- paste0("group_",yr)
  # summarise df_yearly_mun
  df_year <- df_yearly_mun %>%
    dplyr::filter(
      state %in% report_states,
      year == yr) %>%
    dplyr::select(
      all_of(group_col), year,
      pam_area_ha, pam_output_ton,
      ppm_herd, mb_pasture_ha
      ) %>%
    dplyr::group_by(
      across(all_of(group_col)),
      year
      ) %>%
    dplyr::summarise(
      across(everything(), ~sum(.x, na.rm = TRUE)),
      .groups = "drop"
      ) %>%
    dplyr::mutate(
      area_C   = mb_pasture_ha,
      herd_C   = ppm_herd,
      area_S   = pam_area_ha,
      output_S = pam_output_ton
      )
  # calculate yield
  df_year$yield_C <- ifelse(df_year$area_C > 0,
                            df_year$herd_C / df_year$area_C,
                            0)
  df_year$yield_S <- ifelse(df_year$area_S > 0,
                            df_year$output_S/df_year$area_S,
                            0)
  # summarise df_brazil_mun
  df_mun <- df_brazil_mun %>%
    dplyr::filter(
      state %in% report_states
      ) %>%
    dplyr::select(
      all_of(group_col),
      state,
      mun_area_km2,
      mun_protected_area_km2
      ) %>%
    dplyr::group_by(
      across(all_of(group_col)),
      state
      ) %>%
    dplyr::summarise(
      mun_area_km2           = sum(mun_area_km2,           na.rm=TRUE),
      mun_protected_area_km2 = sum(mun_protected_area_km2, na.rm=TRUE)
      ) %>%
    dplyr::mutate(
      mun_protected_area = mun_protected_area_km2 / mun_area_km2
      )
  # join df_year and df_mun
  df_final <- df_mun %>%
    dplyr::left_join(df_year, by=group_col)
  # return summarised data.frame
  return(df_final)
} # close summarise_mun_yield_f(yr)

# summarise by municipality group
mun_1985_yield <- summarise_mun_yield_f("1985") %>% rename(group = group_1985)
mun_1995_yield <- summarise_mun_yield_f("1995") %>% rename(group = group_1995)
mun_2006_yield <- summarise_mun_yield_f("2006") %>% rename(group = group_2006)
mun_2017_yield <- summarise_mun_yield_f("2017") %>% rename(group = group_2017)

## very few 0 values for cattle, it's probably safe to use the minimum productivity as the lower bound
#sum(mun_1985_yield$yield_C==0) # 1
#sum(mun_1995_yield$yield_C==0) # 0
#sum(mun_2006_yield$yield_C==0) # 1
#sum(mun_2017_yield$yield_C==0) # 0
#
## too many 0 values for soybean - need to deal with this somehow
#sum(mun_1985_yield$yield_S==0) # 614
#sum(mun_1995_yield$yield_S==0) # 803
#sum(mun_2006_yield$yield_S==0) # 989
#sum(mun_2017_yield$yield_S==0) # 883
#
## when looking at the whole country, there are a few extreme values in the IBGE data
## when looking at GO, MT, MS, MA, TO, PI, BA, and DF, soy yields from PAM have reasonable values
#summary(df_brazil_mun$FAO_soy_yield)
#summary(mun_1985_yield$yield_S[mun_1985_yield$yield_S>0])
#summary(mun_1995_yield$yield_S[mun_1995_yield$yield_S>0])
#summary(mun_2006_yield$yield_S[mun_2006_yield$yield_S>0])
#summary(mun_2017_yield$yield_S[mun_2017_yield$yield_S>0])

mun_yields_years <- rbind(mun_1985_yield, mun_1995_yield,
                          mun_2006_yield, mun_2017_yield)

# plot dispersion of yield_S (from IBGE's PAM)
plot_box_S_years <- mun_yields_years %>%
  dplyr::filter(area_S > 0) %>%
  ggplot(
    aes(x = factor(year), y = yield_S, fill = factor(year))
  ) +
  geom_boxplot(outlier.size=0.5) +
  scale_fill_manual(values=c("1985"="#7b3b63","1995"="#9e5c84","2006"="#b56d95","2017"="#C27BA0"),
                    guide="none") +
  facet_wrap(~state, ncol=4, scales="free_y") +
  labs(title    = "Soybean yield (tons/ha, IBGE)",
       subtitle = paste0("Realized yield in census years\nMunicipalities with soy area > 0 ha"),
       x="Year", y="tons / ha") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))

# plot dispersion of yield_S (from FAO GAEZ)
# get municipalities with soybean area > 0 ha in 2017
df_mun_with_soy_area_2017 <- df_yearly_mun %>%
  dplyr::filter(
    year          ==   2017,
    mb_soybean_ha >    0,
    state         %in% report_states
  )
mun_with_soy_area_2017 <- df_mun_with_soy_area_2017$code_mun
# create plot
plot_box_S_fao <- df_brazil_mun %>%
  dplyr::filter(code_mun %in% mun_with_soy_area_2017) %>%
  ggplot(
    aes(x = factor(state), y = FAO_soy_yield, fill = factor(state))
  ) +
  geom_boxplot(outlier.size=0.5) +
  labs(title    = "Soybean yield (tons/ha, FAO)",
       subtitle = paste0("Avg. attainable yield in 1981-2010\nMunicipalities with soy area > 0 ha in 2017"),
       x="State", y="tons / ha", fill="State") +
  guides(fill="none") + # omit legent with state names
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))


( plot_box_S_years | plot_box_S_fao )

save_fig(plot_box_S_years, "states_yield_S_box_plot_yearly")
save_fig(plot_box_S_fao,   "states_yield_S_box_plot_fao")

rm(mun_yields, summarise_mun_yield_f,
   df_mun_with_soy_area_2017, mun_with_soy_area_2017,
   mun_1985_yield, mun_1995_yield,
   mun_2006_yield, mun_2017_yield,
   plot_box_S_years, plot_box_S_fao)

```
```{r mun_group_map, include=FALSE}

# associate each municipality in the shapefile with a grouping code
groups_sf <- cities_sf %>%
  dplyr::left_join(
    dplyr::select(df_brazil_mun,
                  code_mun:group_2017, name_biome),
    by = "code_mun"
    )

# create a shapefile for each municipality group and associate the geometries
# with yield_C and yield_S

groups_1985_sf <- groups_sf %>% 
  dplyr::select(group=group_1985, geometry) %>% 
  dplyr::group_by(group) %>%
  dplyr::summarise(geometry=st_union(geometry)) %>%
  dplyr::left_join(
    dplyr::filter(mun_yields_years, year=="1985"),
    by = "group"
  )

groups_1995_sf <- groups_sf %>% 
  dplyr::select(group=group_1995, geometry) %>% 
  dplyr::group_by(group) %>%
  dplyr::summarise(geometry=st_union(geometry)) %>%
  dplyr::left_join(
    dplyr::filter(mun_yields_years, year=="1995"),
    by = "group"
  )

groups_2006_sf <- groups_sf %>% 
  dplyr::select(group=group_2006, geometry) %>% 
  dplyr::group_by(group) %>%
  dplyr::summarise(geometry=st_union(geometry)) %>%
  dplyr::left_join(
    dplyr::filter(mun_yields_years, year=="2006"),
    by = "group"
  )

groups_2017_sf <- groups_sf %>% 
  dplyr::select(group=group_2017, geometry) %>% 
  dplyr::group_by(group) %>%
  dplyr::summarise(geometry=st_union(geometry)) %>%
  dplyr::left_join(
    dplyr::filter(mun_yields_years, year=="2017"),
    by = "group"
  )

rm(groups_sf)
```

```{r mun_yield_S_map, echo=FALSE}

plot_yield_S_1985 <- groups_1985_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_S), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80") +
  labs(title = "Soybean yield (1985)",
       fill  = "tons/ha.")

plot_yield_S_1995 <- groups_1995_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_S), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80") +
  labs(title = "Soybean yield (1995)",
       fill  = "tons/ha.")

plot_yield_S_2006 <- groups_2006_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_S), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80") +
  labs(title = "Soybean yield (2006)",
       fill  = "tons/ha.")

plot_yield_S_2017 <- groups_2017_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_S), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80") +
  labs(title = "Soybean yield (2017)",
       fill  = "tons/ha.")

plot_yield_S_FAO <- cities_sf %>%
  dplyr::left_join(df_brazil_mun, by="code_mun") %>%
  ggplot() +
  geom_sf(aes(fill=FAO_soy_yield), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80") +
  labs(title = "Soybean yield (FAO)",
       fill  = "tons/ha.")

(plot_yield_S_1985 | plot_yield_S_1995)
(plot_yield_S_2006 | plot_yield_S_2017)
plot_yield_S_FAO

save_fig(plot_yield_S_1985, "map_mun_yield_S_1985")
save_fig(plot_yield_S_1995, "map_mun_yield_S_1995")
save_fig(plot_yield_S_2006, "map_mun_yield_S_2006")
save_fig(plot_yield_S_2017, "map_mun_yield_S_2017")
save_fig(plot_yield_S_FAO,  "map_mun_yield_S_FAO")

rm(plot_yield_S_1985, plot_yield_S_1995,
   plot_yield_S_2006, plot_yield_S_2017,
   plot_yield_S_FAO)

```

## Distribution of municipality-level cattle yield

```{r mun_yield_C_distrib, echo=FALSE}
# plot dispersion of yield_C (head of cattle per hectare)
plot_box_C_years <- mun_yields_years %>%
  dplyr::filter(area_C > 0) %>%
  ggplot(
    aes(x = factor(year), y = yield_C, fill = factor(year))
  ) +
  geom_boxplot(outlier.size=0.5) +
  scale_fill_manual(values=c("1985"="#a38b00","1995"="#c9b84a","2006"="#e0cf7a","2017"="#edde8e"),
                    guide="none") +
  facet_wrap(~state, ncol=4, scales="free_y") +
  labs(title    = "Cattle yield (head/hectare) distribution across years",
       subtitle = paste0("Municipalities with pasture area > 0 ha"),
       x="Year", y="tons / ha") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))

# check outliers
plot_yield_C_outliers <- mun_yields_years %>%
  dplyr::filter(
    mb_pasture_ha > 0,
    yield_C > 2
    ) %>%
  ggplot(aes(x     = mun_protected_area,
             y     = yield_C,
             color = state)) +
  facet_wrap(~ year, ncol = 2) +
  geom_point(size = 1.5, alpha = 0.8) +
  labs(title    = "Cattle yield vs protected area share",
       subtitle = "Municipalities with pasture area > 0 ha and cattle/ha > 2",
       x        = "Protected area (%)",
       y        = "Cattle yield (head/ha)",
       color    = "State")

save_fig(plot_box_C_years, "states_yield_C_box_plot_yearly")
save_fig(plot_yield_C_outliers, "mun_yield_C_outliers")

plot_box_C_years
plot_yield_C_outliers

rm(plot_box_C_years, plot_yield_C_outliers)
```


```{r mun_yield_C_map, echo=FALSE}
# on scale_fill_viridis_c() with just arguments `option` and `na.value`, I end up with mostly black maps because of the outliers. There are two ways around it:
# (1) add arguments `limits=c(0,4)` and `oob=scales::squish` to limit the color scale and make all outliers have the same color
# (2) add argument `trans="log10"` to log the scale (useful for showing the full distribution while compressing outliers)


# (1) CAP MAPS
plot_yield_C_1985_cap <- groups_1985_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", limits=c(0,2), oob=scales::squish) +
  labs(title = "Cattle yield (1985, cap)",
       fill  = "head/ha.")

plot_yield_C_1995_cap <- groups_1995_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", limits=c(0,2), oob=scales::squish) +
  labs(title = "Cattle yield (1995, cap)",
       fill  = "head/ha.")

plot_yield_C_2006_cap <- groups_2006_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", limits=c(0,2), oob=scales::squish) +
  labs(title = "Cattle yield (2006, cap)",
       fill  = "head/ha.")

plot_yield_C_2017_cap <- groups_2017_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", limits=c(0,2), oob=scales::squish) +
  labs(title = "Cattle yield (2017, cap)",
       fill  = "head/ha.")

# (2) LOG MAPS
plot_yield_C_1985_log <- groups_1985_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", trans="log10") +
  labs(title = "Cattle yield (1985, log)",
       fill  = "head/ha.")

plot_yield_C_1995_log <- groups_1995_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", trans="log10") +
  labs(title = "Cattle yield (1995, log)",
       fill  = "head/ha.")

plot_yield_C_2006_log <- groups_2006_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", trans="log10") +
  labs(title = "Cattle yield (2006, log)",
       fill  = "head/ha.")

plot_yield_C_2017_log <- groups_2017_sf %>%
  ggplot() +
  geom_sf(aes(fill=yield_C), color=NA) +
  scale_fill_viridis_c(option = "magma", na.value="grey80", trans="log10") +
  labs(title = "Cattle yield (2017, log)",
       fill  = "head/ha.")

# show plots, save figures, remove objects
(plot_yield_C_1985_cap | plot_yield_C_1985_log)
(plot_yield_C_1995_cap | plot_yield_C_1995_log)
(plot_yield_C_2006_cap | plot_yield_C_2006_log)
(plot_yield_C_2017_cap | plot_yield_C_2017_log)

save_fig(plot_yield_C_1985_cap, "map_mun_yield_C_1985_cap")
save_fig(plot_yield_C_1995_cap, "map_mun_yield_C_1995_cap")
save_fig(plot_yield_C_2006_cap, "map_mun_yield_C_2006_cap")
save_fig(plot_yield_C_2017_cap, "map_mun_yield_C_2017_cap")

save_fig(plot_yield_C_1985_log, "map_mun_yield_C_1985_log")
save_fig(plot_yield_C_1995_log, "map_mun_yield_C_1995_log")
save_fig(plot_yield_C_2006_log, "map_mun_yield_C_2006_log")
save_fig(plot_yield_C_2017_log, "map_mun_yield_C_2017_log")

rm(plot_yield_C_1985_cap, plot_yield_C_1985_log,
   plot_yield_C_1995_cap, plot_yield_C_1995_log,
   plot_yield_C_2006_cap, plot_yield_C_2006_log,
   plot_yield_C_2017_cap, plot_yield_C_2017_log)

rm(groups_1985_sf, groups_1995_sf,
   groups_2006_sf, groups_2017_sf,
   mun_yields_years)
```

<!--
## Comparing IBGE's datasets

The PAM and PPM surveys get municipality-level yearly data on crops and cattle farms for 1974-2024 (though I only use data for 1985-2024), while the Agriculture Census gets municipality-level data on all farms about once a decade since 1920 (though I only use data from 1985, 1995, 2006, and 2017). Below, I compare the numbers for soybean yield (in tons per hectare) and cattle yield (in cattle head per hectare) from both data sources for 1995, 2006 and 2017. I also consider Mapbiomas' municipality-level LULC data to check how far it is from official IBGE numbers for pasture area and soybean/crop area.

```{r compare_IBGE, echo=FALSE}

#yearly_data <- df_yearly_mun %>%
#  dplyr::filter(
#    year %in% c(1995,2006,2017)
#    ) %>%
#  dplyr::select(
#    code_mun, group_1995, group_2006, group_2017, code_amc,
#    name_mun, state, name_micro, name_meso, legal_amazon, year,
#    pam_area_ha, pam_output_ton, ppm_herd,
#    mb_pasture_ha, mb_soybean_ha, mb_tcrop_ha, mb_pcrop_ha
#  )
#
#census_data <- df_census_mun %>%
#  dplyr::filter(
#    area_group == "total",
#    day_group == "total"
#  ) %>%
#  dplyr::select(
#    code_mun, year,
#    activity_group,
#    farm_area_ha, n_farms, activity_area
#  )

```
-->

# Land Use and Cover Change
## Brazil
```{r yearly_LULC, echo=FALSE}

# (1) first I generate a plot with all LULC categories from yearly_data
yearly_LULC_data <- df_yearly_mun %>%
  dplyr::select(year, mb_natcover_ha, mb_nonveg_ha, mb_notobs_ha, mb_pasture_ha,
                mb_soybean_ha, mb_tcrop_ha, mb_forestry_ha, mb_pcrop_ha) %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(across(everything(), ~sum(.x, na.rm = TRUE)),
                   .groups = "drop") %>%
  dplyr::mutate(total_area = mb_natcover_ha + mb_nonveg_ha + mb_notobs_ha + mb_pasture_ha +
                  mb_soybean_ha + mb_tcrop_ha + mb_forestry_ha + mb_pcrop_ha) %>%
  dplyr::transmute(year,
                   mb_natcover_ha = 100 * mb_natcover_ha / total_area,
                   mb_nonveg_ha   = 100 * mb_nonveg_ha   / total_area,
                   mb_notobs_ha   = 100 * mb_notobs_ha   / total_area,
                   mb_pasture_ha  = 100 * mb_pasture_ha  / total_area,
                   mb_soybean_ha  = 100 * mb_soybean_ha  / total_area,
                   mb_tcrop_ha    = 100 * mb_tcrop_ha    / total_area,
                   mb_forestry_ha = 100 * mb_forestry_ha / total_area,
                   mb_pcrop_ha    = 100 * mb_pcrop_ha    / total_area)

lulc_order <- c("mb_natcover_ha", "mb_forestry_ha", "mb_pasture_ha",
                "mb_soybean_ha", "mb_tcrop_ha", "mb_pcrop_ha",
                "mb_nonveg_ha", "mb_notobs_ha"
                )

plot_LULC1 <- yearly_LULC_data %>%
  tidyr::pivot_longer(cols=-year, names_to="category",values_to="value") %>%
  dplyr::mutate(category=factor(category,levels=lulc_order)) %>%
  ggplot(aes(x=year, y=value, fill=category)) +
  geom_area() +
 scale_fill_manual(
    values = c(
      "mb_natcover_ha" = "#1f8d49",
      "mb_forestry_ha" = "#7a5900",
      "mb_pasture_ha"  = "#edde8e",
      "mb_soybean_ha"  = "#f5b3c8",
      "mb_tcrop_ha"    = "#C27BA0",
      "mb_pcrop_ha"    = "#d082de",
      "mb_nonveg_ha"   = "#d4271e",
      "mb_notobs_ha"   = "#838B8B"
    ),
    labels = c(
      "mb_natcover_ha" = "Natural cover",
      "mb_forestry_ha" = "Forestry",
      "mb_pasture_ha"  = "Pasture",
      "mb_soybean_ha"  = "Soybean",
      "mb_tcrop_ha"    = "Temporary crops",
      "mb_pcrop_ha"    = "Perennial crops",
      "mb_nonveg_ha"   = "Non-vegetated",
      "mb_notobs_ha"   = "Not observed"
    )
  ) +
  labs(title="Yearly LULC shares",
       x="Year",
       y="Share of total area (%)",
       fill="Land use category") +
  theme_minimal()


# (2) I also generate a simplified LULC plot, containing only the LULC categories present in the model
yearly_LULC2 <- yearly_LULC_data %>%
  dplyr::transmute(year,
                   veg = mb_natcover_ha + mb_forestry_ha,
                   cattle = mb_pasture_ha,
                   crop = mb_soybean_ha + mb_tcrop_ha + mb_pcrop_ha,
                   abandoned = mb_nonveg_ha + mb_notobs_ha)

lulc_order2 <- c("veg", "cattle", "crop", "abandoned")

plot_LULC2 <- yearly_LULC2 %>%
  tidyr::pivot_longer(cols=-year, names_to="category",values_to="value") %>%
  dplyr::mutate(category=factor(category,levels=lulc_order2)) %>%
  ggplot(aes(x=year, y=value, fill=category)) +
  geom_area() +
 scale_fill_manual(
    values = c(
      "veg"       = "#1f8d49",
      "cattle"    = "#edde8e",
      "crop"      = "#C27BA0",
      "abandoned" = "#d4271e"
    ),
    labels = c(
      "veg"       = "Vegetation",
      "cattle"    = "Pasture",
      "crop"      = "Crops",
      "abandoned" = "Non-vegetated"
    )
  ) +
  labs(title="Yearly LULC shares (simplified)",
       x="Year",
       y="Share of total area (%)",
       fill="Land use category") +
  theme_minimal()

# show plots
(plot_LULC1 | plot_LULC2)
save_fig(plot_LULC1,"Brazil_lulc_8classes")
save_fig(plot_LULC2,"Brazil_lulc_4classes")

# clean up
rm(yearly_data2, yearly_LULC_data, yearly_LULC2, lulc_order, lulc_order2, yearly_data, plot_var,
   plot_LULC1, plot_LULC2)

```

## States
```{r yearly_LULC_by_state, echo=FALSE}
# get lulc variables for state-year pairs
lulc_state <- df_yearly_mun %>%
  dplyr::filter(state %in% report_states) %>%
  dplyr::select(state, year, mb_natcover_ha, mb_nonveg_ha, mb_notobs_ha, mb_pasture_ha,
                mb_soybean_ha, mb_tcrop_ha, mb_forestry_ha, mb_pcrop_ha) %>%
  dplyr::group_by(state,year) %>%
  dplyr::summarise(across(everything(), ~sum(.x, na.rm = TRUE)),
                   .groups = "drop") %>%
  dplyr::mutate(total_area = mb_natcover_ha + mb_nonveg_ha + mb_notobs_ha + mb_pasture_ha +
                  mb_soybean_ha + mb_tcrop_ha + mb_forestry_ha + mb_pcrop_ha) %>%
  dplyr::transmute(year, state,
                   veg       = (mb_natcover_ha + mb_forestry_ha)/total_area,
                   cattle    = (mb_pasture_ha)/total_area,
                   crop      = (mb_soybean_ha + mb_tcrop_ha + mb_pcrop_ha)/total_area,
                   abandoned = (mb_nonveg_ha + mb_notobs_ha)/total_area
                   ) %>%
  tidyr::pivot_longer(cols=c(veg,cattle,crop,abandoned),
                      names_to="category", values_to="share") %>%
  dplyr::mutate(category=factor(category, levels=c("veg","cattle","crop","abandoned")))

# plot
plot_lulc_state <- ggplot(lulc_state, aes(x=year, y=share, fill=category)) +
  geom_area() +
  scale_fill_manual(
    values = c("veg"="#1f8d49", "cattle"="#edde8e",
               "crop"="#C27BA0", "abandoned"="#d4271e"),
    labels = c("veg"="Vegetation", "cattle"="Pasture",
               "crop"="Crops",     "abandoned"="Non-vegetated")
  ) +
  scale_y_continuous(labels=scales::percent) +
  facet_wrap(~state, ncol=4) +
  labs(title="Yearly LULC shares by Cerrado state (simplified)",
       x="Year", y="Share of total area", fill="Category") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45, hjust=1))

plot_lulc_state
save_fig(plot_lulc_state,"states_lulc")

rm(lulc_state, plot_lulc_state)

```

# Labor intensity

<!--
Using data from IBGE's Census of Agriculture (available for 1985, 1995, 2006, and 2017) at the state level, I calculate the number of workers per hectare in three sectors of activity: cattle ranching, temporary crops and forestry. Below, I provide the numbers for states with large Cerrado areas and the country as a whole:
-->

```{r labor_intensity_ha, echo=FALSE}

# state-level data
labor_intensity_state <- df_census_state %>%
  dplyr::filter(
    state          %in% report_states,
    area_group     ==   "total",
    activity_group %in% c("cattle", "temp_crop"),
    day_group      ==   "total"
    ) %>%
  dplyr::select(
    state, year, activity_group, 
    activity_area, workers
    ) %>%
  dplyr::mutate(
    workers_per_ha = workers/activity_area
    ) %>%
  dplyr::select(
    state, activity_group, year, workers_per_ha
    ) %>%
  tidyr::pivot_wider(
    names_from  = c(activity_group, year),
    values_from = workers_per_ha
    ) %>%
  dplyr::select(
    state, cattle_1995, temp_crop_1995, cattle_2006, temp_crop_2006, cattle_2017, temp_crop_2017
    )

# national-level data
labor_intensity_national <- df_census_state %>%
  dplyr::filter(
    area_group     ==   "total",
    activity_group %in% c("cattle", "temp_crop"),
    day_group      ==   "total"
    ) %>%
  dplyr::select(
    activity_group, year, activity_area, workers
    ) %>%
  dplyr::group_by(
    activity_group,
    year
    ) %>%
  dplyr::summarise(
    activity_area = sum(activity_area, na.rm=TRUE),
    workers       = sum(workers,       na.rm=TRUE),
    .groups = "drop"
  ) %>%
  dplyr::transmute(
    state = "Brazil",
    activity_group,
    year,
    workers_per_ha = workers/activity_area
  ) %>%
  tidyr::pivot_wider(
    names_from  = c(activity_group, year),
    values_from = workers_per_ha
  ) %>%
  dplyr::select(
    state, cattle_1995, temp_crop_1995, cattle_2006, temp_crop_2006, cattle_2017, temp_crop_2017
  )

# bind and order columns
table_wph <- base::rbind(labor_intensity_state, labor_intensity_national)

# generate table
table_wph %>%
  kableExtra::kable(col.names=c("Area", "Cattle", "Crops", "Cattle", "Crops", "Cattle", "Crops"),
                    digits=4,
                    caption=paste0("Workers per Hectare by Sector")) %>%
  kableExtra::add_header_above(c(" "=1, "1995"=2, "2006"=2, "2017"=2)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width=FALSE) %>%
  kableExtra::row_spec(nrow(table_wph),bold=TRUE)

rm(labor_intensity_state, labor_intensity_national)

```

When working at the 1km resolution, we have

```{r labor_intensity_plots_1km, echo=FALSE}

# get values per 100 hectares (1 sq km)
table_wph[,2:7] <- table_wph[,2:7]*100

# generate table
table_wph %>%
  kableExtra::kable(col.names=c("Area", "Cattle", "Crops", "Cattle", "Crops", "Cattle", "Crops"),
                    digits=4,
                    caption=paste0("Workers per Plot (1km x 1km) by activity")) %>%
  kableExtra::add_header_above(c(" "=1, "1995"=2, "2006"=2, "2017"=2)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width=FALSE) %>%
  kableExtra::row_spec(nrow(table_wph),bold=TRUE)

```

When working at the 5km resolution, we have

```{r labor_intensity_plots_5km, echo=FALSE}

# get values per 2500 hectares (25 sq km)
table_wph[,2:7] <- table_wph[,2:7]*25

# generate table
table_wph %>%
  kableExtra::kable(col.names=c("Area", "Cattle", "Crops", "Cattle", "Crops", "Cattle", "Crops"),
                    digits=4,
                    caption=paste0("Workers per Plot (5km x 5km) by activity")) %>%
  kableExtra::add_header_above(c(" "=1, "1995"=2, "2006"=2, "2017"=2)) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width=FALSE) %>%
  kableExtra::row_spec(nrow(table_wph),bold=TRUE)

rm(table_wph)
```

# Land Concentration

Using data from IBGE's Census of Agriculture (available for 1995, 2006, and 2017) at the state level, I calculate the share of farm area in each of the following area classes:
- less than 100 hectares;
- from 100 to 1000 hectares;
- more than 1000 hectares.
This data is available at the state level, but I also provide results for the whole country.

Since IBGE omits data whenever there are too few farms in an area class, I correct the tables to establish a minimum area threshold for farms in each area class - for instance, if a municipality has one single farm with more than 1000 hectares and its area is registered as zero, I consider this farm to have 1000 hectares. These artificial lower bounds ensure that land concentration figures make sense. This issue only shows up in SE, PE, AL, CE, PB, RN, PR and RJ in 2017, so this shouldn't be a problem for the Cerrado or the Amazon, but this manipulation slightly changes the country-level land concentration measures.

```{r land_concentration_all_years, echo=FALSE}

# get state-level data on farm number and farm area by area group
farm_area <- df_census_state %>%
  dplyr::filter(
    activity_group ==   "total",
    day_group      ==   "total",
    area_group     %in% c("under100ha", "from100to1kha", "over1kha", "total")
  ) %>%
  dplyr::select(
    state, year, area_group,
    farm_area_ha, n_farms
  ) %>%
  dplyr::mutate(
    avg_area = farm_area_ha/n_farms
  )

# fix 'farm_area_ha' for each area group
farm_under100ha    <- farm_area %>% dplyr::filter(area_group=="under100ha")    # no issues
farm_from100to1kha <- farm_area %>% dplyr::filter(area_group=="from100to1kha") # no issues
farm_over1kha      <- farm_area %>% dplyr::filter(area_group=="over1kha") # 8 rows with avg_area lower than 1kha
# the problematic rows are for SE, PE, AL, CE, PB, RN, PR, and RJ in 2017
farm_over1kha$farm_area_ha[farm_over1kha$avg_area<1000] <- 1000*farm_over1kha$n_farms[farm_over1kha$avg_area<1000]
farm_over1kha$avg_area <- farm_over1kha$farm_area_ha / farm_over1kha$n_farms
# get farm_total
farm_total <- farm_area %>% 
  dplyr::filter(area_group=="total") %>%
  dplyr::select(state, year) %>%
  dplyr::left_join(farm_under100ha,    by=join_by(state, year)) %>%
  dplyr::left_join(farm_from100to1kha, by=join_by(state, year)) %>%
  dplyr::left_join(farm_over1kha,      by=join_by(state, year)) %>%
  dplyr::transmute(
    state, year,
    area_group   = "total",
    farm_area_ha = farm_area_ha.x + farm_area_ha.y + farm_area_ha,
    n_farms      = n_farms.x + n_farms.y + n_farms,
    avg_area     = farm_area_ha / n_farms
    )
# join everything back
farm_area <- rbind(
  farm_under100ha,
  farm_from100to1kha,
  farm_over1kha,
  farm_total
  ) %>%
  dplyr::select(-avg_area)
# add a final row for the whole country
farm_area_brazil <- farm_area %>%
  dplyr::select(-state) %>%
  dplyr::group_by(year, area_group) %>%
  dplyr::summarise(farm_area_ha = sum(farm_area_ha, na.rm=TRUE),
                   n_farms      = sum(n_farms, na.rm=TRUE),
                   .groups = "drop") %>%
  dplyr::transmute(state="Brazil", year, area_group, farm_area_ha, n_farms)
# join and clean up
farm_area <- rbind(farm_area, farm_area_brazil)
rm(farm_under100ha, farm_from100to1kha,
   farm_over1kha, farm_total, farm_area_brazil)

# get share of farm areas
area_shares <- farm_area %>%
  dplyr::filter(
    area_group != "total",
    state %in% c(report_states, "Brazil")
    ) %>%
  dplyr::group_by(state, year) %>%
  dplyr::mutate(share_area = 100 * farm_area_ha / sum(farm_area_ha)) %>%
  dplyr::ungroup() %>%
  dplyr::select(state, year, area_group, share_area) %>%
  tidyr::pivot_wider(
    names_from  = c(area_group, year),
    values_from = share_area
  )

# get shares of farm count
farm_shares <- farm_area %>%
  dplyr::filter(
    area_group != "total",
    state %in% c(report_states, "Brazil")
    ) %>%
  dplyr::group_by(state, year) %>%
  dplyr::mutate(share_farms = 100 * n_farms / sum(n_farms)) %>%
  dplyr::ungroup() %>%
  dplyr::select(state, year, area_group, share_farms) %>%
  tidyr::pivot_wider(
    names_from  = c(area_group, year),
    values_from = share_farms
  )

# Table 1)
area_shares %>%
  kable(
    digits  = 1,
    caption = "Share of total farm area (%) by size class",
    col.names = c("State", rep(c("<100", "100–1000", ">1000"), 4))
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "1985" = 3, "1995" = 3, "2006" = 3, "2017" = 3)) %>%
  kableExtra::row_spec(nrow(area_shares),bold=TRUE)

# Table 2)
farm_shares %>%
  kable(
    digits  = 1,
    caption = "Share of farms (%) by size class",
    col.names = c("State", rep(c("<100", "100–1000", ">1000"), 4))
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "1985" = 3, "1995" = 3, "2006" = 3, "2017" = 3)) %>%
  kableExtra::row_spec(nrow(farm_shares),bold=TRUE)


# Plot) stacked bar chart
plot_brazil_land_concentration <- area_shares %>%
  dplyr::filter(state == "Brazil") %>%
  tidyr::pivot_longer(
    cols      = -state,
    names_to  = c("area_group", "year"),
    names_sep = "_",
    values_to = "share_area"
  ) %>%
  dplyr::mutate(
    year = as.numeric(year),
    area_group = factor(area_group,
                        levels = c("under100ha", "from100to1kha","over1kha"),
                        labels = c("<100",       "100-1000",     ">1000"))
  ) %>%
  ggplot(aes(x = factor(year), y = share_area, fill = area_group)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = c("<100"      = "#a8d1e7",
                                "100-1000" = "#2a8db5",
                                ">1000"    = "#0d2d6b")) +
  labs(title = "National farm area by size class",
       x     = "Census year",
       y     = "Share of total farmland area",
       fill  = "Size class") +
  theme_minimal()

gridExtra::grid.arrange(plot_brazil_land_concentration)

save_fig(plot_brazil_land_concentration,"Brazil_land_concentration")

# clean up
rm(farm_area, area_shares, farm_shares, plot_brazil_land_concentration)
  
```

The 2017 Census of Agriculture has more classes of farm size, so we can take a more detailed look in that year:

```{r land_concentration_2017, echo=FALSE}
# get state-level data on farm number and farm area by area group
farm_area <- df_census_state %>%
  dplyr::filter(
    activity_group ==   "total",
    day_group      ==   "total",
    year           ==   2017,
    area_group     %in% c("under100ha", "from100to1kha", "from1kto2500ha", "from2500to10kha", "over10kha", "total")
  ) %>%
  dplyr::select(
    state, area_group,
    farm_area_ha, n_farms
  ) %>%
  dplyr::mutate(
    avg_area = farm_area_ha/n_farms
  )

# fix 'farm_area_ha' for each area group
farm_under100ha      <- farm_area %>% dplyr::filter(area_group=="under100ha")      # no issues
farm_from100to1kha   <- farm_area %>% dplyr::filter(area_group=="from100to1kha")   # no issues
farm_from1kto2500ha  <- farm_area %>% dplyr::filter(area_group=="from1kto2500ha")  #  9 rows with avg_area < 1kha
farm_from2500to10kha <- farm_area %>% dplyr::filter(area_group=="from2500to10kha") # 12 rows with avg_area < 10kha
farm_over10kha       <- farm_area %>% dplyr::filter(area_group=="over10kha")       # 19 rows with avg_area < 10kha
farm_over10kha$avg_area[farm_over10kha$state=="SE"] <- 0
farm_over10kha$avg_area[farm_over10kha$state=="DF"] <- 0
# fix problematic rows
farm_from1kto2500ha$farm_area_ha[farm_from1kto2500ha$avg_area<1000] <- 1000*farm_from1kto2500ha$n_farms[farm_from1kto2500ha$avg_area<1000]
farm_from2500to10kha$farm_area_ha[farm_from2500to10kha$avg_area<2500] <- 2500*farm_from2500to10kha$n_farms[farm_from2500to10kha$avg_area<2500]
farm_over10kha$farm_area_ha[farm_over10kha$avg_area<10000] <- 10000*farm_over10kha$n_farms[farm_over10kha$avg_area<10000]

farm_from1kto2500ha$avg_area  <- farm_from1kto2500ha$farm_area_ha / farm_from1kto2500ha$n_farms
farm_from2500to10kha$avg_area <- farm_from2500to10kha$farm_area_ha / farm_from2500to10kha$n_farms
farm_over10kha$avg_area       <- farm_over10kha$farm_area_ha / farm_over10kha$n_farms

# get farm_total
farm_total <- farm_area %>% 
  dplyr::filter(area_group=="total") %>%
  dplyr::select(state) %>%
  dplyr::left_join(farm_under100ha,      by="state") %>%
  dplyr::left_join(farm_from100to1kha,   by="state") %>%
  dplyr::left_join(farm_from1kto2500ha,  by="state") %>%
  dplyr::left_join(farm_from2500to10kha, by="state") %>%
  dplyr::left_join(farm_over10kha,       by="state") %>%
  dplyr::transmute(
    state,
    area_group   = "total",
    farm_area_ha = farm_area_ha + farm_area_ha.x + farm_area_ha.x.x + farm_area_ha.y + farm_area_ha.y.y,
    n_farms      = n_farms + n_farms.x + n_farms.x.x + n_farms.y + n_farms.y.y,
    avg_area     = farm_area_ha / n_farms
    )
# join everything back
farm_area <- rbind(
  farm_under100ha,
  farm_from100to1kha,
  farm_from1kto2500ha,
  farm_from2500to10kha,
  farm_over10kha,
  farm_total
  ) %>%
  dplyr::select(-avg_area)
# add a final row for the whole country
farm_area_brazil <- farm_area %>%
  dplyr::select(-state) %>%
  dplyr::group_by(area_group) %>%
  dplyr::summarise(farm_area_ha = sum(farm_area_ha, na.rm=TRUE),
                   n_farms      = sum(n_farms, na.rm=TRUE),
                   .groups = "drop") %>%
  dplyr::transmute(state="Brazil", area_group, farm_area_ha, n_farms)
# join and clean up
farm_area <- rbind(farm_area, farm_area_brazil)
rm(farm_under100ha, farm_from100to1kha, farm_from1kto2500ha,
  farm_from2500to10kha, farm_over10kha, farm_total, farm_area_brazil)

# get share of farm areas
area_shares <- farm_area %>%
  dplyr::filter(
    area_group != "total",
    state %in% c(report_states, "Brazil")
    ) %>%
  dplyr::group_by(state) %>%
  dplyr::mutate(share_area = 100 * farm_area_ha / sum(farm_area_ha)) %>%
  dplyr::ungroup() %>%
  dplyr::select(state, area_group, share_area) %>%
  tidyr::pivot_wider(
    names_from  = area_group,
    values_from = share_area
  )

# get shares of farm count
farm_shares <- farm_area %>%
  dplyr::filter(
    area_group != "total",
    state %in% c(report_states, "Brazil")
    ) %>%
  dplyr::group_by(state) %>%
  dplyr::mutate(share_farms = 100 * n_farms / sum(n_farms)) %>%
  dplyr::ungroup() %>%
  dplyr::select(state, area_group, share_farms) %>%
  tidyr::pivot_wider(
    names_from  = area_group,
    values_from = share_farms
  )

# Table 1)
area_shares %>%
  kable(
    digits  = 1,
    caption = "Share of total farm area (%) by size class, 2017",
    col.names = c("State", "<100", "100-1000", "1000-2500", "2500-10000", ">10000")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::row_spec(nrow(area_shares),bold=TRUE)

# Table 2)
farm_shares %>%
  kable(
    digits  = 1,
    caption = "Share of farms (%) by size class, 2017",
    col.names = c("State", "<100", "100-1000", "1000-2500", "2500-10000", ">10000")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  kableExtra::row_spec(nrow(farm_shares),bold=TRUE)

# clean up
rm(farm_area, area_shares, farm_shares)

```



```{r land_concentration_parameters, echo=FALSE}

#We can use this data to get rough estimates for the values of parameters `share_smallprod`, `share_mediumprod`, `share_largeprod`, `ratio_medium_to_small`, and `ratio_large_to_small`. If the cell grid has 1km resolution, each cell has 100 hectares and we can initialize the model with values from 1985, 1995, 2006 or 2017 with no issues. If the cell grid has 5km resolution, each cell has 2500 hectares and only the 2017 data has a 1-1 relation with AgriCerrado's grid structure - we can still use the data from 1985, 1995 or 2006, but we must be aware that this is a gross approximation.

# helper function to compute the 5 parameters for a given set of states and year
# small  = <100ha
# medium = 100-1000ha
# large  = >1000ha

#get_params_3class <- function(states, yr) {
#  farm_number <- farm_number_by_class %>%
#    dplyr::filter(state %in% states, year==yr) %>%
#    dplyr::summarise(small  = sum(less_than_100_ha,    na.rm=TRUE),
#                     medium = sum(from_100_to_1000_ha, na.rm=TRUE),
#                     large  = sum(more_than_1000_ha,   na.rm=TRUE),
#                     total  = sum(total,               na.rm=TRUE))
#  
#  farm_area <- farm_area_by_class %>%
#    dplyr::filter(state %in% states, year==yr) %>%
#    dplyr::summarise(small  = sum(less_than_100_ha,    na.rm=TRUE),
#                     medium = sum(from_100_to_1000_ha, na.rm=TRUE),
#                     large  = sum(more_than_1000_ha,   na.rm=TRUE))
#  
#  data.frame(share_small  = 100 * farm_number$small  / farm_number$total,
#             share_medium = 100 * farm_number$medium / farm_number$total,
#             share_large  = 100 * farm_number$large  / farm_number$total,
#             avg_small    = farm_area$small  / farm_number$small,
#             avg_medium   = farm_area$medium / farm_number$medium,
#             avg_large    = farm_area$large  / farm_number$large)
#  
#} # close get_params_3class()
#
## helper function to compute the 5 parameters for a given set of states and year
## small  = <2500ha
## medium = 2500-10000ha
## large  = >10000ha
#get_params_5class <- function(states) {
#  farm_number <- farm_table_2017 %>%
#    dplyr::filter(state %in% states) %>%
#    dplyr::summarise(small  = sum(less_than_100_ha, na.rm=TRUE) +
#                              sum(from_100_to_1000_ha, na.rm=TRUE) +
#                              sum(from_1000_to_2500_ha, na.rm=TRUE),
#                     medium = sum(from_2500_to_10000_ha, na.rm=TRUE),
#                     large  = sum(more_than_10000_ha, na.rm=TRUE),
#                     total  = sum(total, na.rm=TRUE))
#  
#  farm_area <- area_table_2017 %>%
#    dplyr::filter(state %in% states) %>%
#    dplyr::summarise(small  = sum(less_than_100_ha, na.rm=TRUE) +
#                              sum(from_100_to_1000_ha, na.rm=TRUE) +
#                              sum(from_1000_to_2500_ha, na.rm=TRUE),
#                     medium = sum(from_2500_to_10000_ha, na.rm=TRUE),
#                     large  = sum(more_than_10000_ha, na.rm=TRUE))
#  
#  data.frame(share_small  = 100 * farm_number$small  / farm_number$total,
#             share_medium = 100 * farm_number$medium / farm_number$total,
#             share_large  = 100 * farm_number$large  / farm_number$total,
#             avg_small    = farm_area$small  / farm_number$small,
#             avg_medium   = farm_area$medium / farm_number$medium,
#             avg_large    = farm_area$large  / farm_number$large)
#    
#} # closes get_params_5class()
#
## Table 1) Land Concentration (3 classes - 1985, 1995, 2006 and 2017)
## build wide table
#years_3class <- c("1985", "1995", "2006", "2017")
## process data
#rows_3class <- lapply(c(cerrado_states, "Brazil"), function(s) {
#  states <- if (s == "Brazil") unique(farm_number_by_class$state) else s
#  year_cols <- lapply(years_3class, function(yr) {
#    df <- get_params_3class(states, yr)
#    # rename columns to include year suffix before cbind
#    names(df) <- paste0(names(df), "_", yr)
#    df
#  })
#  result <- base::do.call(cbind, year_cols)
#  result$area <- if (s == "Brazil") "Brazil (national)" else s
#  result <- result %>% dplyr::select(area, dplyr::everything())
#  result
#})
## create final df
#table_3class <- base::do.call(base::rbind, rows_3class)
## generate table for report
## Table 1a) Land Concentration (3 classes - 1985 and 1995)
#table_3class %>%
#  dplyr::select(area,
#                ends_with("_1985"),
#                ends_with("_1995")) %>%
#  kableExtra::kable(
#    col.names = c("Area",
#                  rep(c("Small (%)", "Medium (%)", "Large (%)",
#                        "Avg. small (ha)", "Avg. medium (ha)", "Avg. large (ha)"), 2)),
#    digits  = 1,
#    caption = "Land concentration parameters (small <100 ha, medium 100–1000 ha, large >1000 ha) — part 1/2"
#  ) %>%
#  kableExtra::add_header_above(c(" " = 1, "1985" = 6, "1995" = 6)) %>%
#  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"),
#                            full_width = FALSE) %>%
#  kableExtra::row_spec(nrow(table_3class), bold = TRUE)
#
## Table 1b) Land Concentration (3 classes - 2006 and 2017)
#table_3class %>%
#  dplyr::select(area,
#                ends_with("_2006"),
#                ends_with("_2017")) %>%
#  kableExtra::kable(
#    col.names = c("Area",
#                  rep(c("Small (%)", "Medium (%)", "Large (%)",
#                        "Avg. small (ha)", "Avg. medium (ha)", "Avg. large (ha)"), 2)),
#    digits  = 1,
#    caption = "Land concentration parameters (small <100 ha, medium 100–1000 ha, large >1000 ha) — part 2/2"
#  ) %>%
#  kableExtra::add_header_above(c(" " = 1, "2006" = 6, "2017" = 6)) %>%
#  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"),
#                            full_width = FALSE) %>%
#  kableExtra::row_spec(nrow(table_3class), bold = TRUE)
#
##table_3class %>%
##  kableExtra::kable(col.names = c("Area",
##                                  rep(c("Small (%)", "Medium (%)", "Large (%)", "Avg. small (ha)", "Avg. medium (ha)", #"Avg. #large (ha)"),4)),
##                    digits  = 1,
##                    caption = "Land concentration parameters (small <100 ha, medium 100–1000 ha, large >1000 ha)") %>%
##  add_header_above(c(" " = 1,
##                     "1985" = 6, "1995" = 6,
##                     "2006" = 6, "2017" = 6)) %>%
##  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
##  row_spec(nrow(table_3class), bold = TRUE)
#
## Table 2) Land Concentration (3 classes - 2017)
## process data
#rows_5class <- lapply(c(cerrado_states, "Brazil"), function(s) {
#  states <- if (s == "Brazil") unique(farm_number_by_class_2017$state) else s
#  get_params_5class(states) %>%
#    dplyr::mutate(area = if (s == "Brazil") "Brazil (national)" else s,
#                  .before = 1)
#})
## create final df
#table_5class <- base::do.call(base::rbind, rows_5class)
## generate table for report
#table_5class %>%
#  kableExtra::kable(col.names=c("Area", "Small (%)", "Medium (%)", "Large (%)", "Avg. small (ha)", "Avg. medium (ha)", "Avg. #large (ha)"),
#                    digits=1,
#                    caption="Land concentration parameters (small <2500 ha, medium 2500–10000 ha, large >10000 ha)") %>%
#  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
#                            full_width = FALSE) %>%
#  kableExtra::row_spec(nrow(table_5class), bold=TRUE)
#
## clean up
#rm(get_params_3class, years_3class, rows_3class, table_3class,
#   get_params_5class, rows_5class, table_5class,
#   area_table_2017, farm_table_2017,
#   size_class_labels)

```

# Area by activity type
Using data from IBGE's Census of Agriculture (available for 1995, 2006, and 2017) at the state level, I calculate the share of farm area in each of the following classes of economic activity:
- cattle ranching;
- crop farming;
- forestry;
- other.
This data is available at the state level, but I also provide results for the whole country

```{r area_by_activity_type, echo=FALSE}

# get state-level area by activity
area_by_activity <- df_census_state %>%
  dplyr::filter(
    area_group ==   "total",
    day_group  ==   "total",
    year       %in% c(1995, 2006, 2017)
    ) %>%
  dplyr::select(
    state, year, activity_group, activity_area
  ) %>%
  tidyr::pivot_wider(
    names_from  = activity_group,
    values_from = activity_area
  ) %>%
  dplyr::mutate(
    crop  = perm_crop + temp_crop,
    total = cattle + crop + forestry
  ) 

# get country-level area by activity
area_by_activity_brazil <- area_by_activity %>%
  dplyr::select(-state) %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(
    total     = sum(total, na.rm=TRUE),
    cattle    = sum(cattle, na.rm=TRUE),
    forestry  = sum(forestry, na.rm=TRUE),
    perm_crop = sum(perm_crop, na.rm=TRUE),
    temp_crop = sum(temp_crop, na.rm=TRUE),
    crop      = sum(crop, na.rm=TRUE)
  ) %>%
  dplyr::transmute(
    state = "Brazil",
    year, total, cattle, forestry, perm_crop, temp_crop, crop
  )

# join in a single table
area_by_activity <- rbind(area_by_activity, area_by_activity_brazil)
rm(area_by_activity_brazil)

# generate final table
area_by_activity <- area_by_activity %>%
  dplyr::transmute(
    state, 
    year,
    cattle   = 100*cattle   / total,
    crop     = 100*crop     / total,
    forestry = 100*forestry / total
  ) %>%
  tidyr::pivot_longer(
    cols      = cattle:forestry,
    names_to  = "activity_class",
    values_to = "value"
  ) %>%
  tidyr::pivot_wider(
    names_from  = c(activity_class, year),
    values_from = value
  ) %>%
  dplyr::filter(
    state %in% c(report_states, "Brazil")
  )
  
# Table) area shares
area_by_activity %>%
  kableExtra::kable(
    digits = 1,
    caption="Share of total farm area (%) by activity class",
    col.names=c("State", rep(c("Cattle", "Crop", "Forestry"), 3)),
    ) %>%
  kableExtra::add_header_above(c(" "=1, "1995"=3, "2006"=3, "2017"=3)) %>%
  kableExtra::kable_styling(bootstrap_options = c("stripend", "hover", "condensed"),
                            full_width=FALSE) %>%
  kableExtra::row_spec(nrow(area_by_activity),bold=TRUE)

# Plot) stacked bar chart
# national area shares by size class (19985, 1995, 2006, 2017)
plot_activity_area <- area_by_activity %>%
  dplyr::filter(state=="Brazil") %>%
  tidyr::pivot_longer(
    cols = -state,
    names_to  = c("activity_class","year"),
    names_sep = "_",
    values_to = "share"
    ) %>%
  dplyr::mutate(
    year = as.numeric(year),
    activity_class = factor(activity_class,
                            levels = c("cattle", "crop", "forestry"),
                            labels = c("Cattle", "Crop", "Forestry"))
    ) %>%
  ggplot(aes(x=factor(year), y=share, fill=activity_class)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(scale=1)) +
  scale_fill_manual(values = c("Cattle"   = "#edde8e",
                               "Crop"     = "#C27BA0",
                               "Forestry" = "#1f8d49")) +
  labs(title   = "National farm area by activity class",
       x       = "Census year",
       y       = "Share of total farmland area",
       fill    = "Activity class") +
  theme_minimal()

gridExtra::grid.arrange(plot_activity_area)

save_fig(plot_activity_area,"Brazil_farmland_by_activity_census")

# clean up
rm(area_by_activity, plot_activity_area)

```



